---
title: "Police Shooting Fatalities in the United States"
author: "Neha Gyawali"
date: "4/26/2022"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
## Introduction
Unnecessary violence employed by many police officers throughout the US is an issue that needs to be addressed and dealt with. Recently, police fatalities have been highlighted in the media for their often unjust natures. Black men especially seem to be more targeted than the rest of the population. Let's take a look at police fatalities data that has been gathered by the Washington Post starting from 2015 to now to help us understand what demographic is at risk. We will take a look at the top 25 most fatal cities and break down the fatalities by race. We will also compare the race breakdown of the fatalities to the race breakdown of the population of the cities.
---
## Package Installation
```{r}
#install.packages("usmap")

# Load tidyverse to manipulate data
# Load usmap to make plot
# Load ggplot2 and plotly for graphing,

library(tidyverse, warn.conflicts = F)
options(tidyverse.quiet = TRUE)
library(usmap, warn.conflicts = F)
library(ggplot2, warn.conflicts = F)
library(plotly, warn.conflicts = F)

#read data
data <- read_csv("fatal-police-shootings-data.csv")
top25 <- read_csv("top25.csv")
top25 <- top25[1:25,]
top25 <- top25[,1:8]
```

## Gender Breakdown
Here we can see the difference in the number of female versus male fatalities. This helps us understand the demographic that is in danger of being killed by a police officer.
```{r}

genderBreakdown <- data %>% group_by(gender) %>%  summarise(count = n())
genderBreakdown


gender <- c(327, 6913)
color =  c("mistyrose", "lightblue")
pie( gender, labels = c("4.5%", "95.48%"), col = color)
legend("topleft", legend = c("Female", "Male"),
       fill =  c("mistyrose", "lightblue"))
title("Gender Breakdown of Police Fatalities")



```

## Police Fatalities 

Some text about this table and how interesting the results are! 

```{r}
lat_long <- data %>% select(longitude, latitude, race, name, date, manner_of_death, armed, age, city, flee) %>% rename(lat = latitude, lon = longitude)
lat_long <- na.omit(lat_long)

transformed_lat_long <- usmap_transform(lat_long)

p <- plot_usmap(regions = "states") + geom_point(data = transformed_lat_long, aes(x = x, y=y, color =race, text = paste("Name: ", name, "\nArmed: ", armed, "\nAge: ", age, "\nCity: ", city, "\nManner of Death: ", manner_of_death, "\nDate: ", date, "\nFleeing: ", flee )), alpha = .05, position = "jitter")  + scale_color_manual(values = c("Asian" = "navyblue",  "Hispanic" = "cadetblue4", "Native American" = "darkolivegreen", "Other" = "darkviolet", "White"= "royalblue", "Black" = "red4"))+ labs(title = "Police Fatalities Locations")  + theme(panel.background=element_blank()) 

p <- ggplotly(p) 
p
```

## Race Breakdown of Police Fatalities in the top 25 Most Fatal Cities

```{r}
#find cities with most fatalities
numberfatalities <- data %>% group_by(city) %>%  summarise(count = n()) %>%  arrange(desc(count))


#take top 30 most fatal cities
mostDeaths <- numberfatalities[1:25,]

topdeadliestBreakdown <- merge(data, mostDeaths)


#graph with race breakdown
d <- ggplot(data = topdeadliestBreakdown, aes(x = reorder(city, city, function(x)-length(x)), fill = race)) + geom_bar() + theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5)) + ggtitle("Race Breakdown with Fatality Count") + ylab("Fatality Count") + xlab("City") + labs(fill = "Race")  +scale_fill_manual(values = c("Asian" = "navyblue",  "Hispanic" = "cadetblue4", "Native American" = "darkolivegreen", "Other" = "darkviolet", "White"= "royalblue", "Black" = "red4", "NA" = "lavenderblush4"))
d
```

```{r}
d <- ggplot(data = topdeadliestBreakdown, aes(x = reorder(city, city, function(x)-length(x)), fill = race)) + geom_bar(position = "fill") + theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5)) + labs(title = "Race Breakdown with Proportions", fill = "Race")+ xlab("City") + ylab("Proportion") +scale_fill_manual(values = c("Asian" = "navyblue",  "Hispanic" = "cadetblue4", "Native American" = "darkolivegreen", "Other" = "darkviolet", "White"= "royalblue", "Black" = "red4", "NA" = "lavenderblush4"))
d
```
Using these graphs we can see the top 25 cities with the most police fatalities. The first graph shows us the race breakdown in numbers and the second one shows us the race breakdown in percentage.

```{r}

```
---
```{r}
data4 <- data  %>% mutate(date = (as.Date(date, "%m/%d/%Y"))) %>%  group_by(date)%>% mutate(count = n())
data4


m <- ggplot(data4, aes(x=date, y=count)) +geom_line()
m
```
 
